---
title: "Patterns in Support for Reparations - Survey Analysis"
output: html_document
---

<style type="text/css">

h1 {
  text-align: center;
  font-weight: bold;
  font-family: 'Helvetica Neue';
}

h2 {
 font-family: 'Helvetica Neue';
}

body{
 font-family: 'Helvetica Neue';
 font-size: 16px;
}

</style>

<br>
<br>



```{r setup, include=FALSE}
# for many data tricks!
library(tidyverse)
# for a few handy personal functions
library(willbprocessed)
# easy calculations on survey data
library(srvyr)
# for pretty tables
library(kableExtra)
# for paths
library(here)
# color palletes
library(viridis)

```


```{r colors, include=FALSE}
# Colors for plots
# same colors as first post for consistency
v_pal <- viridis::viridis(5)
v_pal8 <- viridis::viridis(8)
# new colors for reparations support
magma_pal6 <- viridis::magma(6)
magma_pal4 <- viridis::magma(4)
# creating smaller palletes from those ^
# generations plots
gen3_colors <- c(v_pal[3], v_pal[2], v_pal[1])
gen6_colors <- c(v_pal8[6], v_pal8[5], v_pal8[4], v_pal8[3], v_pal8[2], v_pal8[1])
# reparations plots
rep5_colors <- c(magma_pal6[2],magma_pal6[3], magma_pal6[6], magma_pal6[5], magma_pal6[4])
rep3_colors <- c(magma_pal4[2], magma_pal4[4], magma_pal4[3])


```

```{r functions, include=FALSE}
# the font for this post
post_font <- "Helvetica Neue"

# decreasing copy-paste for similar pies
one_var_pie <- function(df, varname, color_pal){
  ggplot(data = df, aes(x="", fill = .data[[varname]], y = pct)) + 
  geom_bar(stat = "identity", color = "black") +
  geom_text(
    aes(label = Percent, x = 1.6, y = pct),
    size = 5,
    position = position_stack(vjust = 0.5),
    fontface = "bold",
    color = "black") +
  # to make it pie chart
  coord_polar(theta = "y") +
  scale_fill_manual(values = color_pal) +
  # no axes
  theme_void() +
  theme(legend.position = "right",
        legend.title = element_blank(),
        plot.title = element_text(size = 20, hjust = 0.5, face = "bold"),
        text = element_text(family = post_font))
  
}
# THIS IS GOOD, IMPLIMENT TIME
format_plot_data <- function(df, dummies, varname, var_levels, group_vars = c()){
  df %>%
    # select only cols we need, if no group_vars provided it will just do id and weight
    select(all_of(c("respondent_id", "weight", group_vars))) %>%
    bind_cols(dummies) %>%
    pivot_longer(
      # patterns based on our use cases. not robust lol
      cols = matches(" |Support|Oppose|Agree|Disagree"),
      names_to = varname,
      values_to = "value") %>%
    # levels so right orders in plots
    mutate(!!varname := factor(.data[[varname]], levels = var_levels)) %>%
    # thanks for making this easy srvyr!
    srvyr::as_survey(weights = weight, ids = respondent_id) %>%
    srvyr::group_by_at(c(group_vars, varname)) %>%
    srvyr::summarise(pct = survey_mean(value, vartype = "ci", na.rm = T)) %>%
  mutate(
    # round * 100
    across(all_of(contains("pct")), willbprocessed::as_percent, 0, F),
    # for labels
    Percent = paste0(pct, "%"),
  )
}

# creating table's for reparations breakdown
hr40_tabs <- function(srvyr_df){
  srvyr_df %>%
  srvyr::summarise(
    Yes = survey_mean(I(aware_hr40 == "Yes"), vartype = "ci", na.rm = T),
    No = survey_mean(I(aware_hr40 == "No"), vartype = "ci", na.rm = T)
    ) %>%
    #for now cuz such big error bars
    mutate(across(all_of(matches("Yes|No")), as_percent, 0, T)) %>% 
    select(-all_of(matches("upp|low")))
}


```


```{r clean_dataset, include=FALSE}
# data is not public
df <- readRDS(here("Not For Git/Clean_Data/ras_20210615.RDS")) %>%
  mutate(
    # the weight for this dataset
    weight = weight_rep_unique,
    # going from 5 categories to 3
    reparations3 = str_extract(reparations,"support|oppose|Not sure") %>% str_to_sentence() %>%
      factor(levels = c("Support", "Oppose", "Not sure")),
    # generations 3 cat
    generations3 = 
      recode(as.character(generations), "Strongly agree" = "Agree", "Somewhat agree" = "Agree", 
                        "Neither agree nor disagree" = "Other", "Somewhat disagree" = "Disagree", 
                        "Strongly disagree" = "Disagree", 
                        "Prefer not to say" = "Other") %>%
      factor(levels = c("Agree", "Other", "Disagree"))
  ) %>%
  # NAs are those that do not have responses to reparations question
  filter(!is.na(weight))
# dummies for what level people support reparations
rep_dummies <- factor_to_dummies(df$reparations)
# dummies for what support, oppose, not sure
rep3_dummies <- factor_to_dummies(df$reparations3)
# dummies for generations categories
gen_dummies <- factor_to_dummies(df$generations)

# srvyr version of the df
srvyr_df <-  srvyr::as_survey(df, weights = weight, ids = respondent_id)

```

About half of Americans either support or are not sure if they support the US government giving financial compensation to the descendants of enslaved people. While a majority of white people oppose reparations, a majority of Black Americans and a thin plurality of Hispanic people support them. Furthermore a majority of Americans who acknowledge lasting effects of slavery and discrimination on Black Americans support reparations, while those that deny this reality oppose reparations by an overwhelming majority. Together, these patterns suggest that the appetite for reparations is significant and significant portions of the population are open to the financial compensation of enslaved people.
<br>

## Disclaimer

The data for this project comes from a multi-wave survey project I conducted while at YouGov. When I left, I was allowed to take the data, graphics and write-ups I had worked on to further analyze the data and post it here on my own website. <i>All of the views expressed in this write-up are my own</i>. They do not reflect the views or opinions of YouGov.
<br>

## Half of Americans Are Open To Reparations

While at YouGov, I fielded three waves of a national survey between December 2020 and June 2021 to gauge Americans’ awareness of systemic racism and level of support for policies that aim to promote racial equity.

This sample has a combined N of a unique 2,700 respondents. It was weighted to be nationally representative on age, gender, race, education, census region, voter registration, news interest, and 2016 presidential vote. Weights ranged from 0.03 to 6.02, with a mean of 1, SD of 0.83 and a MOE of +/- 2.45 percent. All graphics and tables incorporate these weights.

In this survey, I asked respondents about their view on reparations for slavery. I chose to explicitly gauge support for reparations given to the descendents of enslaved people rather than asking about reparations for both slavery and centuries of discriminatory policy. Were I to run this again, I would like to adjust the wording to see how such a change would affect the patterns of support discussed in the remainder of this post.
<br>

```
Would you support the United States giving financial compensation to the individual descendants of enslaved people?
 <1> Strongly support
 <2> Somewhat support
 <3> Somewhat oppose
 <4> Strongly oppose
 <5> Not sure
```
<br>

```{r reparations_processing, include=FALSE}

reparations_breakdown <- format_plot_data(
  df, 
  dummies = rep_dummies,
  varname = "reparations",
  var_levels = c("Strongly support", "Somewhat support", "Not sure",
    "Somewhat oppose", "Strongly oppose")
  )


```

```{r reparations_pie, echo=F, fig.width=10, fig.height=7.5, message=F}

one_var_pie(reparations_breakdown, 'reparations', rep5_colors) +
  labs(title = "Support for Reparations", x = "", y = "")

```

Overall, about half of Americans oppose reparations, while a little over 31 percent support them and 18 percent are not sure. The large level of uncertainty is likely in part due to the question’s design. I was attempting to ascertain general attitudes towards reparations for slavery as a concept rather than to measure support for a specific plan or for the term “Reparations” itself. Perhaps, the large level of uncertainty suggests that a significant percent of Americans are open to reparations but would need more details as to how they would be structured. And while a plurality of people do oppose reparations, about half of Americans are not opposed to them. And a closer look at the data reveals some illuminating patterns.
<br>

## Pluralities of Black and Hispanic People Support Reparations

Patterns of support for reparations differ significantly across racial groups. Overall, a plurality of about 43 percent of People of Color support reparations, while a majority of 61 percent of white people oppose them. Within People of Color, sixty percent of Black Americans support reparations, and only 19 percent oppose them. Support among Hispanic Americans is more split. A thin plurality supports reparations, but about equally as many oppose them or are not sure about them. And unfortunately, the confidence intervals on these values are fairly large.
<br>

```{r rep3_race3_processing, include = FALSE}
# format the data. similar to above. should probs fxn this
rep3_race3 <-format_plot_data(
  df, 
  dummies = rep3_dummies,
  varname = "reparations",
  var_levels = c("Not sure", "Oppose", "Support"),
  group_vars = "race4"
  ) %>%
  # error bars are pretty huge
  filter(race4 != "Other") %>%
  # just so things make a little more sense
  rename("race3" = "race4") %>%
  # reverse alphabetical so it will show in Alphabetical
  mutate(race3 = factor(race3, levels = c("White", "Hispanic", "Black")))


```


```{r rep3_race3_bar, fig.width=10, fig.height=7.5, echo=FALSE}
ggplot(data = rep3_race3) +
  # the parameters for the bar plot
  geom_bar(
    aes(x = race3, y = pct, fill = reparations),
    color = "black",
    stat = "identity",
    position = position_dodge(width = 0.9)
    ) +
  # making it width wise
  coord_flip() +
  # the paramters for the text
  geom_text(
    # if I want to put it at the end
    # aes(x = race3, y = pct + 5, label = Percent, group = reparations),
    aes(x = race3, y = (pct/2), label = Percent, group = reparations),
    #position = position_stack(vjust = 0.5),
    position = position_dodge(width = 0.9),
    color = "black",
    fontface = "bold",
 
  ) +
  # now the error bars
  geom_errorbar(
    aes(x = race3, ymin = pct_low, ymax = pct_upp, group = reparations),
    width = 0.2,
    position = position_dodge(width = 0.9)
    ) +
  # colors for the bars
  scale_fill_manual(values = rep3_colors,  breaks = c("Support", "Not sure","Oppose")) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  # theme
  theme_minimal() +
  theme(legend.position = "top",
        legend.title = element_blank(),
        plot.title = element_text(size = 20, hjust = 0.5, face = "bold"),
        axis.text.x= element_blank(),
        text = element_text(family = "Helvetica Neue")) +
  labs(title = "Reparations Support by Race", x = "", y = "")

# let's play with plotly displays perhaps. 
# ^I was gonna do some interactive plotly graphics with some significant differences
# but I think this is good for now.

# Last thing! re order some levels here so alphabetical and legend makes more sense


```

There is insufficient data to determine how these patterns manifest among other groups of People of Color, but this data suggests that white Americans' majority opposition to reparations is somewhat anomalous and is a significant factor in explaining why half of Americans oppose reparations.

It is not a coincidence that those who have benefited most from America’s biased institutions are most likely to oppose actions that attempt to begin to make amends for it’s country’s history of oppression. Perhaps many white respondents are not aware of the ramifications of slavery and the degree to which the systems born out of it still influence life in America. Perhaps others are aware of these realities, but do not believe reparations, as described in the question, would be helpful. And finally some may not wish to attempt to address the lasting impacts of slavery and racial discrimination at all. It is difficult to know respondent’s rationals, but responses to a question aimed to measure awareness the effects of slavery and discrimination provide us with some clues.

## Reparations are Supported by a Majority of those with an Understanding of the Generational Effects of Slavery and Discrimination

In the same survey, respondents were asked to indicate the degree to which they agreed with the following statement: 

```
Generations of slavery and discrimination have created conditions that make it difficult for African Americans to achieve economic mobility
```
<br>

```{r include=FALSE, generations_processing}

generations_breakdown <- format_plot_data(
  df, 
  dummies = gen_dummies,
  varname = "generations",
  var_levels = 
    c("Strongly agree", "Somewhat agree", "Prefer not to say",
      "Neither agree nor disagree", "Somewhat disagree", "Strongly disagree")
  ) 


```

```{r generations_pie, echo=F, fig.width=10, fig.height=7.5, message=F}

one_var_pie(generations_breakdown, 'generations', gen6_colors) +
  labs(title = "Agreement with Generations Statement", x = "", y = "")

```

A plurality of Americans (about 43 percent) agree with the statement, and about 37 percent disagree with it. The majority of those who agree with the statement (55 percent) indicated support for reparations. Alternatively, a vast majority of those who disagree with the statement on the generational effects of slavery and discrimination (87 percent) oppose reparations. Only 25 percent of those who agree with the generational challenges of slavery oppose reperations, while 20 percent remain unsure.
<br>

```{r include=FALSE}

gen3_rep3 <- format_plot_data(
  df, 
  dummies = rep3_dummies,
  varname = "reparations",
  var_levels = c("Not sure", "Oppose", "Support"),
  group_vars = "generations3"
  ) %>%
  mutate(
    reparations = factor(as.character(reparations), levels = c("Support", "Not sure", "Oppose")),
    pct_label = round(pct),
    #for better spacing in plot labels
    pct_label = paste0("   ", pct_label, "%")
  )


```

```{r echo=F, fig.width=10, fig.height=7.5, message=F}


ggplot(data = gen3_rep3) +
  geom_bar(stat = "identity", aes(x = reparations, y = pct, fill = generations3),
           position = "dodge") +
  scale_fill_manual(values = gen3_colors) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  geom_errorbar(aes(x = reparations, ymin = pct_low, ymax = pct_upp, group = generations3),
                width = 0.2, position = position_dodge(width = 0.9)) +
  geom_text(
    aes(x = reparations, y = pct_upp + 5, label = pct_label, group = generations3),
    position = position_dodge(width = 0.9),
    size = 5,
    fontface = "bold",
    color = "black"
    ) +
  theme_minimal() +
  theme(legend.position = "top",
        legend.title = element_blank(),
        plot.title = element_text(size = 20, hjust = 0.5, face = "bold"),
        text = element_text(family = "Helvetica Neue")) +
  labs(title = "Support for Reparations By Generations Statement", x = "", y = "")


```

I believe this data reveals three patterns: (1) There is significant support for reparations for slavery among those who acknowledge the generational effects of slavery and discrimination on Black Americans. (2) Those who refuse to acknowledge the generational effects of slavery and discrimination oppose reparations in great numbers. There is very little uncertainty among this group, even though the description of compensation used in the question was vague. (3) Many of those who do not disagree with the generations statement are not sure as to whether or not they support reparations in the manner described in the question.

I think that most Americans who are able to admit that slavery has lead to generations of sytemic discrimination are ready to consider reparations and unpack what they might look like. Scholars and activists are offering concrete plans and there is a bill in Congress to establish a Commision to look into reparations called H.R. 40. Unfortunately, relatively few people are aware of it.
<br>

```{r hr40_table, echo=FALSE}
# first broken down by reparations support
hr40_bkdwn <- srvyr::group_by(srvyr_df, reparations3) %>%
  hr40_tabs() %>%
  rename("Support for Reparations" = reparations3)
# then overall
hr40_all <- hr40_tabs(srvyr_df) %>%
  mutate(`Support for Reparations` = "All")

# and then together as one html table
hr40_bkdwn %>%
  bind_rows(hr40_all) %>%
  kableExtra::kable(
    caption = "<center><strong>Awareness of HR40 by Support for Reparations</strong></center>",
    align = c("l", "c", "c"),
    table.attr = "style='width:70%;'"
    ) %>%
  add_header_above(c(" " = 1, "Aware of HR40" = 2)) %>%
  kableExtra::kable_paper(html_font = "Helvetica Neue")

```

The survey asked panelists if they were aware of HR40 and only 16 percent were. Furthermore, only about 23 percent of respondents who indicated that they support reparations in our survey also indicated they were aware of HR40 and only 14 percent of those not sure if they support reparations were aware of it. Awareness does seem to be growing overtime, but there is clearly more of an appetite for reparations than the attention they receive from our government and media.

## Looking Forward

I hope that this project helps reinforce the reality that there is a significant appetite for reparations. I believe the patterns discussed in this post are compelling and I hope that posts and discussion like this can help focus dialogue on reparations and hopefully apply pressure to pass HR 40 and make reparations for slavery and disicrimination a priority.


