---
title: "Japanese American Mass Incareration During World War II"
output: html_document
---

<style type="text/css">

h1 {
  text-align: center;
  font-weight: bold;
  font-family: 'Helvetica Neue';
}

h2 {
 font-family: 'Helvetica Neue';
}

body{
 font-family: 'Helvetica Neue';
 font-size: 16px;
}

</style>

<br>
<br>

```{r setup, include=FALSE}
library(tmap)
library(tidyverse)
```

During World War II, the United States Government Incarcerated more than 120,000 Japanese Americans in concentration camps. The history of racism and xenophobia that lead to the creation of these camps and the conditions of life within are seldom discussed or taught, even in California where the majority of Japanese Americans sent to these camps were from.

The purpose of this post is to raise awareness about the camps and utilize incomplete data from the National Archives to help readers visualize the scale of this injustice. This post will not give an exhaustive history of the shameful actions perpetrated by actors in American government and FDR administration to actualize these concentration camps. I encourage the reader to visit <a href="https://densho.org/">densho.org</a> as a first step to learning more. There are also memoirs of the experiences of people in the camps and books such as _Infamy: The Shocking Story of the Japanese American Internment in World War II_ that aim to articulate an overview of what happened.

## The Data

The bulk of the data I acquired for this study comes from the US National Archives. I have posted cleaned data files on <a href="https://github.com/wbannick">my github</a>. The data has a little over 109,000 rows. Therefore, <b>the data is incomplete</b>. Please keep that in mind when exploring the subsequent analysis.

I also created a map using this data. I acquired a map of 1940 county boundaries from <a href="www.nhgis.org">IPUMS NHGIS, University of Minnesota</a>. For my maps I only used data from California, Oregon, and Washington. This data represents 106931 cases or almost 98% of the available data, but there were also Japanese Americans incarcerated from other states, particularly Hawaii and Arizona.

I merged counties to sub-regions identified in the documentation for the data from the National Archives. I further grouped together some sub-regions in which few incarcerated Japanese Americans lived. This allows me to present more reliable summary statistics, but also is somewhat misleading. For example there were relatively fewer people near incarcerated from outside of what is known as the "Exclusion Zone." For the boundaries of this zone as well as other helpful data please explore <a href="https://maps.densho.org/sitesofshame/">Densho's Sites of Shame map</a>. When exploring my map please keep in mind that within the sub-regions presented, fewer people were interned outside the Exclusion Zone.

The map below is colored to reflect the number of Japanese Americans who were incarcerated by the sub-region of their prior residence. Click on any subregion to see more summary statistics for those who had lived in each location.


```{r map_output, echo = FALSE, message = FALSE, warning = FALSE, out.width="100%", out.height= "700px"}
# some data prep so labels look nicer
sf_sub <- sf::read_sf(here::here("Not for Git/Clean_Data/pacific_subregion.gpkg")) %>%
  # recoding names of columns used in this visualization!
  # keeping others as is so that grouped data is easier to program with
  rename(
    # eh may need a better label
    "People Incarcerated" = "n",
    "Percent of Records" = "percent_records",
    "Percent Never Been to Japan" = "never_in_japan",
    "Percent Born in the US" = "us_born",
    "Percent Children" = "under_18",
    "Most Common Camp" = "mode_camp"
  )

tmap::tmap_mode("view")

# creating the map
tmap::tm_shape(sf_sub, name = "People Incarcerated") + 
  tm_polygons(
    "People Incarcerated",
    id = "subregion",
    palette = "-magma",
    popup.vars = 
      c(
        "People Incarcerated",
        "Percent of Records",
        "Percent Never Been to Japan",
        "Percent Born in the US",
        "Percent Children",
        "Most Common Camp"
      ),
    breaks = c(50, 1000, 4000, 10000, 20000, 40000)
    ) +
  tm_view(set.zoom.limits = c(5,10)) +
  tmap_options(basemaps = c("Esri.WorldGrayCanvas")) 

```

## Takeaways

Most of the Japanese Americans sent to Concentration Camps lived in California or in the Seattle and Portland Metropolitan Areas. Almost 84% of the people in the dataset with prior residence data were from California. Another 12% were from Seattle or Portland. All Japanese Americans, including American citizens were targeted for the mass removal so this likely largely reflects where the largest populations of Japanese Americans were within the Exclusion Zone. But these trends are revealing because they illuminate how many lives were uprooted the extent to which regions and communities were effected by this forced removal. 

For example when, I look at the sub-regions in the Bay Area and Central Coast, I can't help but wonder how thee places would be different if the tens of thousands of Japanese Americans who were forcibly expelled from the homes in the 1940s were able to remain, retain their property and build generational wealth. The mass incarceration of Japanese Americans has had a profoundly racist effect on the development of the communities in which many of us have lived in California.

Most Japanese Americans in recorded in the dataset were born in the US and over half had never even been to Japan. There is no excuse to imprison an entire group of people based on their race and ethnicity regardless of where they were born, and this analysis will not imply otherwise. This data is a helpful reminder of the composition of the Japanse Americans incaracerated. It is important to remember that in the subset of records represented by this data, that over 47% had never been to a country that they were imprisoned for association with.



