---
title: "Japanese American Mass Incareration During World War II"
output: html_document
---

<style type="text/css">

h1 {
  text-align: center;
  font-weight: bold;
  font-family: 'Helvetica Neue';
}

h2 {
 font-family: 'Helvetica Neue';
}

body{
 font-family: 'Helvetica Neue';
 font-size: 16px;
}

</style>

<br>
<br>

```{r setup, include=FALSE}
library(tmap)
library(tidyverse)
```

During World War II, the United States Government incarcerated more than 120,000 Japanese Americans in concentration camps. The history of racism and xenophobia that lead to the creation of these camps and the conditions of life within them is seldom taught in schools or discussed.

The purpose of this post is to raise awareness about the camps and utilize incomplete data from the National Archives to help readers visualize the scale of this injustice. This post will not give an exhaustive history of the shameful actions perpetrated by actors in American government and the FDR administration that actualized these concentration camps. But this history is crucial to learn and I encourage the reader to visit <a href="https://densho.org/">densho.org</a> as a first step. There are also memoirs of the experiences of people in the camps and books such as _Infamy: The Shocking Story of the Japanese American Internment in World War II_ that aim to articulate an overview of what happened.

## The Data

The bulk of the data I acquired for this study comes from the US National Archives. I have posted cleaned data files on <a href="https://github.com/wbannick">my github</a>. The data has a little over 109,000 rows. As over 120,000 Japanese Americans were incarcerated <b>the data is incomplete</b>. Please keep that in mind when exploring the subsequent analysis. More details on missingness in this data can be found in the data documentation from the National Archives, which can also be found in my github.

Below is a map I created using this data as well as spatial data of 1940 county boundaries from <a href="www.nhgis.org">IPUMS NHGIS, University of Minnesota</a>. In my map, I used data from California, Oregon, and Washington. This represents 106,931 cases or almost 98% of the available data. There were also Japanese Americans incarcerated from other states, particularly from Hawaii and Arizona.

I merged counties to sub-regions identified in the documentation for the data from the National Archives. I further grouped together some sub-regions in which few incarcerated Japanese Americans had lived. This allowed me to present more reliable summary statistics, but the final presentation is somewhat misleading. For example, there were relatively few people incarcerated from outside of what is known as the "Exclusion Zone." For the boundaries of this zone as well as a more in depth exploration of related spatial data please visit <a href="https://maps.densho.org/sitesofshame/">Densho's Sites of Shame map</a>. When exploring my map, please keep in mind that peoople were not evenly distributed within the sub-regions presented. In particular, fewer people were forcibly removed from areas outside of the Exclusion Zone.

The map below is colored to reflect the number of Japanese Americans who were incarcerated by the sub-region of their prior residence. Click on any subregion to view summary statistics for those who had lived in each location.


```{r map_output, echo = FALSE, message = FALSE, warning = FALSE, out.width="100%", out.height= "700px"}
# some data prep so labels look nicer
sf_sub <- sf::read_sf(here::here("Not for Git/Clean_Data/pacific_subregion.gpkg")) %>%
  # recoding names of columns used in this visualization!
  # keeping others as is so that grouped data is easier to program with
  rename(
    # eh may need a better label
    "People Incarcerated" = "n",
    "Percent of Records" = "percent_records",
    "Percent Never Been to Japan" = "never_in_japan",
    "Percent Born in the US" = "us_born",
    "Percent Children" = "under_18",
    "Most Common Camp" = "mode_camp"
  )

tmap::tmap_mode("view")

# creating the map
tmap::tm_shape(sf_sub, name = "People Incarcerated") + 
  tm_polygons(
    "People Incarcerated",
    id = "subregion",
    palette = "-magma",
    popup.vars = 
      c(
        "People Incarcerated",
        "Percent of Records",
        "Percent Never Been to Japan",
        "Percent Born in the US",
        "Percent Children",
        "Most Common Camp"
      ),
    breaks = c(50, 1000, 4000, 10000, 20000, 40000)
    ) +
  tm_view(set.zoom.limits = c(5,10)) +
  tmap_options(basemaps = c("Esri.WorldGrayCanvas")) 

```

## Takeaways

Most of the Japanese Americans sent to concentration camps lived in California or in the Seattle and Portland metropolitan areas. Almost 84% of the people in the dataset with prior residence data were from California. Another 12% were from Seattle or Portland. All Japanese Americans, including American citizens were targeted by the mass removal and incarceration, so these trends likely reflect where the largest populations of Japanese Americans were within the Exclusion Zone. They thereby give us a better idea of the extent to which regions and communities were affected by this forced removal. 

One is left to wonder how areas like Los Angeles would be different if the tens of thousands of Japanese Americans who were forcibly expelled from their homes in the 1940s were able to remain, retain their property, and build generational wealth. The mass incarceration of Japanese Americans has had a profoundly racist effect on the development of many communities, particularly in California. I believe this map is a sobering reminder of that reality

Other trends stood out in the data. For example, most Japanese Americans in the dataset (over 65%) were born in the United States and almost half (over 47%) had never even been to Japan. Also, about 30% of the records in the data correspond to children (individuals younger than 18 as of 1942). Again this data is incomplete, and it is impossible to know how representative it is of the entire population of Japanese Americans sent to concentration camps, but the presence of tens of thousands of children in this dataset is a sobering reminder of how many young lives were uprooted and forever changed.

## Looking Forward

I hope this brief post can help readers visualize the scale of the racist mass incarceration of Japanese Americans during World War II. I encourage those who engage with it to take the responsibility to do more research on the topic. Once again, I think densho.org is a great place to start.

I also hope people reflect upon the ways in which racist systems of mass incarceration continue to manifest and be justified and excused in this country, including in areas labeled progressive. I hope this reflection includes an exploration of our problematic prison systems as well as the current concentration camps holding migrants from Latin America. This country has a long history of weaponizing incarceration to achieve racist outcomes, it continues to shape communities we live in.
